Incentives are the biggest barriers to widespread data sharing

### 2021-08-11 new-paper.Rmd
small changes to improve readability, a couple of tweaked sentences and moving the superscript citations around so they fit the sentences better.

```{=tex}
\begin{singlespace}
There are numerous arguments that strongly support practice of open science, with societal and individual benefits. For individual researchers, sharing research artifacts such as data can improve trust and transparency, reproducibility of one's own work, and catalyze new collaborations. Despite the general appreciation of the benefits of data sharing, research data are often only available to original investigators. For data that are shared, lack of useful metadata and documentation make them challenging to reuse. In this paper we argue that incentives and infrastructure for making data useful are the biggest barrier to creating a culture of widespread data sharing. We compare data to code, and computational environments in the context of reproducible research and provide some practical guidance on how one can improve the chances of their data being reusable and partially bridge the incentive gap. Other papers have focussed on many general rules for sharing data. This paper focusses on common sense ideas for sharing tabular data, for a target audience of academics working in data science adjacent fields who are about to submit for publication.

@@

Meaningful progress on most computational reproducibility, the ability of an independent researcher to recreate figures, tables and key findings, became more of a practical reality in the late 2000s as new tools and services came into existence. Scientists started sharing more code and software, partly due to the rapid increase in training opportunities made possible by efforts like The Carpentries, combined with the growing adoption of GitHub by scientists [@Ram2013; @Zagalsky2015]. The bigger driver for this is more likely due to the growing popularity of data science as a discipline distinct from statistics [@Donoho2017]. This rapid growth in data science has been catalyzed by a Cambrianesque explosion of open source software tools. Programming languages such as Python, R and Julia have helped scientists implement and share new methods to work with data. Each of these languages enjoys support of thriving communities of researchers and software developers who contribute many of the building blocks that make them popular. Modern scientists now have access to many interactive computational frameworks (e.g. Jupyter, RMarkdown, Stencila, Observable) that can interface with kernels from dozens of programming languages. Beyond notebooks to run the code, a reader will also require access to the same software dependencies. In a study of applied computational research, less than 20% of software described in publications was able to be installed and run [@Collberg2014]. In addition to locating software, one still has to locate the right versions. Dependency management is a controversial topic in programming communities with no easy solutions. While languages like JavaScript have solved this with package managers like npm, it is much harder to accomplish in others like Python and R which are currently supported by a patchwork of solutions. Although this problem is far from solved, technologies like Docker have alleviated some of these concerns. Researchers can share Dockerfiles to recreate computational environments [@Boettiger2015]. Services like Binder can then leverage Docker and make it possible to share the computational environments in which these notebooks were run. The barriers to verification are now low enough that reproducibility is sometimes a click away, free from the misery of software installation [@Jupyter2018].

@@

Nosek [@nosek_2021] describes the strategies for culture change, especially when the status quo is too entrenched. Beyond motivation, skills, tools, and incentives are critical to enable such change. In a five step process, Nosek describes how achieving lasting culture change requires us to show proof of concept that an activity is possible, then improve the infrastructure to make it easy, followed by increased awareness to make it normative. Once there is a critical mass, find ways to reward those who put it into practice. After these steps, it would then be possible to require this of everyone Figure 1). This framework provides a useful lens through which to examine adoption of code, computational environments and data for reproducibility. With code, we have slowly progressed through these stages. While few journals require that authors share code, the venues to share, along with the necessary training have paved the way for incentives and credit. All of the necessary foundation to make code sharing required are in place. For computational environments, Docker makes it possible to share them in most common use cases [@Boettiger2015]. But running them still requires expertise and time on the part of reviewers. More recent open services like Binder [@Jupyter2018] make it possible to run notebooks atop Docker images, but the technology is still new and experiencing growing pains. There is still a long way to go before sharing computational environments in academia becomes mainstream. Finally, for data, we have implemented these steps out of order. The FAIR data principles were introduced in 2016 with good intentions [@Wilkinson2016] and describe the desired attributes for research data. That they be findable (good metadata), accessible (deposited somewhere persistent), interoperable (easily combined with other datasets), and be reusable. We as a community started out by making data required before demonstrating how this would directly benefit those who share data. Researchers are asked to deposit their data in one of many institutional or commercial repositories. Many journals even require it as a condition for publication via policy. However, the cost of preparing data for reuse remains high, and the career incentives to do so are almost non-existent. Until we make data reuse easier and more rewarding, no level of policy is going to improve the state of data sharing.

Despite the increase in data being shared alongside publications, the status quo of data sharing in science is not good. [@Rowhani-Farid2016] showed that 7 of 157 (4.5%) sampled papers in the British Medical Journal shared data, and only 14% of articles claiming open sharing of data actually shared their data. Similarly, [@Stodden2018] showed only 36% of papers in the journal *Science* shared data, despite the Journal's data sharing policy. [@Colavizza2020] reflects similar findings for PLOS and BMC journals. There are numerous examples of similar problems in other fields, such as psychology [@vanpaemel2015] and biology [@schofield2009]. The reality is that while data sharing is increasing in many fields such as ecology and genomics, the amount of data that can be reused is likely lower. To make shared data useful, we need to realign incentives for researchers.

@@

We believe a key audience could be better served by the current literature on data sharing: researchers who have finished data analysis on tabular data, are ready to submit for publication.
Many papers provide rules and guidelines for sharing data, such as providing metadata, citation, licenses, and lists of data repositories to use. These papers provide general advice from the ethical considerations of sharing data [@Moles2013], how to share many different data types (images, video, etc) [@Hart2016], to “Reward Colleagues Who Share Their Data Properly” [@Goodman2014]. There is also specific advice like the importance of very clean and properly shaped data [@White2013], and to use open source software with changes tracked [@Boland2017]. We walk a fine line between providing highly specific advice (e.g. how to specify NULL values) and the overly general (e.g. sharing data is good). Our goal is to provide practical advice for the researchers who are familiar with modern data science tools and develop their work like research compendia.

@@

To make data more reusable, we need to make it easier to generate metadata, standardise good practices within and across communities, and make reusable data sharing a recognized, normalised, part of research. Here we present a set of minimal set of practices to facilitate this change. These include documentation (Section \@ref(readme)), licensing (Section \@ref(licenses)), citation (Section \@ref(cite)), raw and analysis ready data (Section \@ref(rawdata)), code (Section \@ref(scripts)), and sharing in packages (Section \@ref(pkgs)). We also discuss considerations for sharing of data of various sizes (Section \@ref(size)).

We limit the scope of this paper to small to medium sized tabular data as it covers a vast majority of use cases that researchers encounter day to day. For larger datasets that involve complex workflows (e.g., bioinformatics and medical imaging), the kinds of challenges they face are different and accordingly deserve to be written up separately.

@@

Our advice here is not meant to be exhaustive by any means and are insufficient to fully bridge the gap, but provide a starting point.

@@

Use a README file with your data project to communicate its purpose and contents. A README is often the first place users will look to learn more about a project. The README is a concise text file containing guidance on how to read and interpret the directory - the who, what, when, where, and how (<https://betta.io/blog/2017/02/07/developer-experience-github-readmes/> and. For example, a README in a data repository might explain which directories contain data, tidying scripts, and also why data was collected. The README should be placed in the top level of the project, and there can be multiple README files in a project. In the context of datasets, READMEs are particularly useful when there are no reliable standards. Although data repositories contain a large number of datasets associated with peer reviewed papers, not all are usable (example: https://datadryad.org/stash/dataset/doi:10.5061/dryad.rn6k1).  Some of the more recent submissions contain usage notes and a README, which considerably improve reuse potential (e.g. https://datadryad.org/stash/dataset/doi:10.5061/dryad.3xsj3txbz). \

The date of use, and versions of software should be recorded in the README, so users can in the future can replicate similar computing environment or computing conditions. A more ideal approach would be to clearly describe the computing environment (e.g., a Dockerfile) and packages used (e.g., requirements.txt).

@@

We recommend the CC0 as it provides simple rules for attribution, warranty, and use, and is very common. If data that is CC0 is used in scientific publications, you need to cite where you got the data from. There might be some concern that it is possible to conduct fraud, but it will be found out eventually if the original data is archived properly.

@@

Use metadata with your data. As stated at the start of the paper, we refer to tabular data. There are three benefits to a dataset having metadata: (1) Provide context by describing variables to humans; (2) Avoid errors by describing and preserving variable types for machine readability; (3) Make data easier to find online by facilitating indexing [@google-data-search]. Three pieces of metadata should be considered: (1) **variable names**, (2) **variable descriptions**, and (3) **unit text**. These should be curated by a person. Metadata created should be clearly labelled in a folder or in files named METADATA.

@@

A simple metadata file that describes the variable names, variable description, and unit text is a great starting point, and is sometimes known as a codebook. Metadata can be encoded into more formal plain text formats that follow a set standard, such as EML or SCHEMA-LD. A deep understanding of these formats isn't required to use them. For more details, see [@Broman2017; @Ellis2017; @Arslan2019 @Leipzig2021].

@@

One critical advantage of depositing data in a trusted repository \@ref(avenues) is that it comes with many benefits such as preservation, indexing,  discoverability and most importantly, citation with persistent identifier. Authors can take an extra step by including their own CITATION file to give credit to a paper where the data and associated methods are described in detail. When citing data, you should cite the preferred CITATION file in the first instance, else, if there is a data publication, (for example in _Scientific Data_), cite this, and if none of those apply, cite the repository where the data is deposited (e.g., a Zenodo repository).

@@

Although it is fine to share data among collaborators using temporary cloud storage (like Dropbox, Google Drive), once the paper is public, the data needs to be hosted somewhere persistent. While this approach is useful while iterating on the paper, it does not scale to a larger readership and is highly prone to disappearing (lack of permissions, etc.). Providing back up data is particularly relevant in light of the case in [@Schleunes2020; @Viglione2020], where data was suspected of being tampered. The use of raw data allowed for issues to be identified, the absence of raw data not being available in other publications created difficulty in verifying the validity of the data.

Journals provide data papers, where they provide a formal publication explaining data that is archived. This is a familiar avenue for researchers to receive credit and citation. Writing a data paper can be very useful for researchers, however, data papers lack guidelines on how to structure data for effective reuse and sharing. Partner with Dryad to link archived data with research papers. Dryad will also provide professionals to curate your data, thereby improving usability.

@@

Data can be distributed alongside code using a format such as a package in R, Python, and Julia. A package or module structure provides an advantage as it can be "installed" and loaded just like code. Despite the advantages of sharing data as a software package, there are a few major disadvantages. Firstly, packaging data inside a software package makes the data availability language centric. Users who work in other languages are unlikely to download and export data out of a package. Secondly, data size can be a limitation depending on the repository (CRAN, the R repository, for example, limits package sizes to 5Mb although there are workarounds). Other fields, such as bioinformatics have a process of sharing data through repositories like Bioconductor, with larger data size. Thirdly, there is the cost of authors and data curators/custodians understanding intricacies of package development in order to share data. Nevertheless, the frictionless approach to sharing data in software packages makes reuse substantially faster than retrieving and correctly parsing files from a repository. We recommend the software approach only as an additive to proper data curation.

Packaged data and archival on venues like Zenodo are not mutually exclusive. First, archive data in a long-term data repository (Section \@ref(avenues)), before including it in a package. The act of packaging data in software can shave off hours required to prepare it for use. The canocical citation can be baked into the package's citation file.

@@

Small data has more options in how it is shared - you can bundle it with your analysis code (as in \@ref(rawdata)), or in software \@ref(pkgs). For medium data, consider a venue like Zenodo for archiving. Sharing medium data might require it to be broken into separate chunks for reading (e.g., `teaching-1.csv`, `teaching-2.csv`, etc.), and scripts provided to read it. Large data will require liaising with services from \@ref(avenues), or local research institute library to make special arrangements, to ensure there is some way to gain access, and also maintain a DOI.

@@

Although data sharing is widely mandated by funders, journals, and academic societies, the amount of usable data remains appallingly low. The mere act of depositing data does not make it reusable or useful to a potential consumer. Although data citations do not currently impact promotion and tenure packets, there is anecdotal evidence that papers sharing usable data accrue more citations [@Piwowar2007]. Over time, the more useful we make the data, it is more likely incentive mechanisms will take hold.

>>>>

```{=tex}
\begin{singlespace}
There are numerous arguments that strongly support the practice of open science, with societal and individual benefits. For individual researchers, sharing research artifacts such as data can improve trust and transparency, reproducibility of one's own work, and catalyze new collaborations. Despite the general appreciation of the benefits of data sharing, research data are often only available to the original investigators. For data that are shared, lack of useful metadata and documentation make them challenging to reuse. In this paper we argue that incentives and infrastructure for making data useful are the biggest barrier to creating a culture of widespread data sharing. We compare data to code, and computational environments in the context of reproducible research and provide some practical guidance on how one can improve the chances of their data being reusable and partially bridge the incentive gap. Other papers have focussed on many general rules for sharing data. This paper focusses on common sense ideas for sharing tabular data, for a target audience of academics working in data science adjacent fields who are about to submit for publication.

@@

Meaningful progress on most computational reproducibility, the ability of an independent researcher to recreate figures, tables and key findings, became more of a practical reality in the late 2000s as new tools and services came into existence. Scientists started sharing more code and software, partly due to the rapid increase in training opportunities made possible by efforts like The Carpentries, combined with the growing adoption of GitHub by scientists [@Ram2013; @Zagalsky2015]. The bigger driver for this is more likely due to the growing popularity of data science as a discipline distinct from statistics [@Donoho2017]. This rapid growth in data science has been catalyzed by a Cambrianesque explosion of open source software tools. Programming languages such as Python, R and Julia have helped scientists implement and share new methods to work with data. Each of these languages enjoys the support of thriving communities of researchers and software developers who contribute many of the building blocks that make them popular. Modern scientists now have access to many interactive computational frameworks (e.g. Jupyter, RMarkdown, Stencila, Observable) that can interface with kernels from dozens of programming languages. Beyond notebooks to run the code, a reader will also require access to the same software dependencies. In a study of applied computational research, less than 20% of software described in publications was able to be installed and run [@Collberg2014]. In addition to locating software, one still has to locate the right versions. Dependency management is a controversial topic in programming communities with no easy solutions. While languages like JavaScript have solved this with package managers like npm, it is much harder to accomplish in others like Python and R which are currently supported by a patchwork of solutions. Although this problem is far from solved, technologies like Docker have alleviated some of these concerns. Researchers can share Dockerfiles to recreate computational environments [@Boettiger2015]. Services like Binder can then leverage Docker and make it possible to share the computational environments in which these notebooks were run. The barriers to verification are now low enough that reproducibility is sometimes a click away, free from the misery of software installation [@Jupyter2018].

@@

Nosek [@nosek_2021] describes the strategies for culture change, especially when the status quo is too entrenched. Beyond motivation, skills, tools, and incentives are critical to enable such change. In a five step process, Nosek describes how achieving lasting culture change requires us to show proof of concept that an activity is possible, then improve the infrastructure to make it easy, followed by increased awareness to make it normative. Once there is a critical mass, find ways to reward those who put it into practice. After these steps, it would then be possible to require this of everyone (Figure \@ref(fig:incentable-table)). This framework provides a useful lens through which to examine adoption of code, computational environments, and data for reproducibility. With code, we have slowly progressed through these stages. While few journals require that authors share code, the venues to share, along with the necessary training have paved the way for incentives and credit. All of the necessary foundation to make code sharing required are in place. For computational environments, Docker makes it possible to share them in most common use cases [@Boettiger2015]. But running them still requires expertise and time on the part of reviewers. More recent open services like Binder [@Jupyter2018] make it possible to run notebooks atop Docker images, but the technology is still new and experiencing growing pains. There is still a long way to go before sharing computational environments in academia becomes mainstream. Finally, for data, we have implemented these steps out of order. The FAIR data principles were introduced in 2016 with good intentions [@Wilkinson2016] and describe the desired attributes for research data. That they be findable (good metadata), accessible (deposited somewhere persistent), interoperable (easily combined with other datasets), and be reusable. We as a community started out by making data required before demonstrating how this would directly benefit those who share data. Researchers are asked to deposit their data in one of many institutional or commercial repositories. Many journals even require it as a condition for publication via policy. However, the cost of preparing data for reuse remains high, and the career incentives to do so are almost non-existent. Until we make data reuse easier and more rewarding, no level of policy is going to improve the state of data sharing.

Despite the increase in data being shared alongside publications, the status quo of data sharing in science is not good. Rowhani-Farid et al. [@Rowhani-Farid2016] showed that 7 of 157 (4.5%) sampled papers in the British Medical Journal shared data, and only 14% of articles claiming open sharing of data actually shared their data. Similarly, Stodden et al. [@Stodden2018] showed only 36% of papers in the journal *Science* shared data, despite the Journal's data sharing policy. Similar findings have been found for PLOS and BMC journals [@Colavizza2020]. There are numerous examples of similar problems in other fields, such as psychology [@vanpaemel2015] and biology [@schofield2009]. The reality is that while data sharing is increasing in many fields such as ecology and genomics, the amount of data that can be reused is likely lower. To make shared data useful, we need to realign incentives for researchers.

@@

We believe a key audience could be better served by the current literature on data sharing: researchers who have finished data analysis on tabular data, and are ready to submit for publication.
Many papers provide rules and guidelines for sharing data, such as providing metadata, citation, licenses, and lists of data repositories to use. These papers provide general advice from the ethical considerations of sharing data [@Moles2013], how to share many different data types (images, video, etc) [@Hart2016], to “Reward Colleagues Who Share Their Data Properly” [@Goodman2014]. There is also specific advice like the importance of very clean and properly shaped data [@White2013], and to use open source software with changes tracked [@Boland2017]. We walk a fine line between providing highly specific advice (e.g. how to specify NULL values) and the overly general (e.g. sharing data is good). Our goal is to provide practical advice for the researchers who are familiar with modern data science tools and develop their work like research compendia.

@@

To make data more reusable, we need to make it easier to generate metadata, standardise good practices within and across communities, and make reusable data sharing a recognized, normalised, part of research. Here we present a set of common sense practices to facilitate this change. These include documentation (Section \@ref(readme)), licensing (Section \@ref(licenses)), metadata (Section \@ref(metadata)), raw and analysis ready data (Section \@ref(rawdata)), code (Section \@ref(scripts)), avenues to share data (Section \@ref(avenues)) and citation (Section \@ref(cite)). We also discuss considerations for sharing in packages (Section \@ref(pkgs)) and sharing of data of various sizes (Section \@ref(size)).

We limit the scope of this paper to small to medium sized tabular data as it covers a vast majority of use cases that researchers encounter day to day. Larger datasets with complex workflows (e.g., bioinformatics and medical imaging), face different kinds of challenges and accordingly deserve to be written up separately.

@@

Our advice here is not meant to be exhaustive by any means and is insufficient to fully bridge the gap, but instaed, to provide a starting point.

@@

Use a README file with your data project to communicate its purpose and contents. A README is often the first place users will look to learn more about a project. The README is a concise text file containing guidance on how to read and interpret the directory - the who, what, when, where, and how (<https://betta.io/blog/2017/02/07/developer-experience-github-readmes/>). For example, a README in a data repository might explain which directories contain data, tidying scripts, and also why data was collected. The README should be placed in the top level of the project, and there can be multiple README files in a project (e.g., in different directories). In the context of datasets, READMEs are particularly useful when there are no reliable standards. Although data repositories contain a large number of datasets associated with peer reviewed papers, not all are usable (example: https://datadryad.org/stash/dataset/doi:10.5061/dryad.rn6k1).  Some of the more recent submissions contain usage notes and a README, which considerably improve reuse potential (e.g. https://datadryad.org/stash/dataset/doi:10.5061/dryad.3xsj3txbz). \

The date of use, associated paper, and versions of software should be recorded in the README, so users can in the future can replicate similar computing environment or computing conditions. A more ideal approach would be to clearly describe the computing environment (e.g., a Dockerfile) and packages used (e.g., requirements.txt).

@@

We recommend the CC0 as it provides simple rules for attribution, warranty, and use, and is very common. If data that is CC0 is used in scientific publications, you need to cite where you got the data from as part of scientific practice. There might be some concern that it is possible to conduct fraud, but it will be found out eventually if the original data is archived properly.

@@

Use metadata with your data. There are three benefits to a dataset having metadata: (1) Provide context by describing variables to humans; (2) Avoid errors by describing and preserving variable types for machine readability; (3) Make data easier to find online by facilitating indexing [@google-data-search]. Three pieces of metadata should be considered: (1) **variable names**, (2) **variable descriptions**, and (3) **unit text**. These should be curated by a person. Metadata created should be clearly labelled in a folder or in files named METADATA.

@@

A simple metadata file that describes the variable names, variable description, and unit text is a great starting point, and is sometimes known as a codebook. Metadata can be encoded into more formal plain text formats that follow a set standard, such as EML or SCHEMA-LD. A deep understanding of these formats isn't required to use them. For more details, see [@Broman2017; @Ellis2017; @Arslan2019; @Leipzig2021].

@@

@@

Although it is fine to share data among collaborators using temporary cloud storage (like Dropbox, Google Drive), once the paper is public, the data needs to be hosted somewhere persistent. While this approach is useful while iterating on the paper, it does not scale to a larger readership and is highly prone to disappearing (lack of permissions, etc.). Providing back up data is particularly relevant in light of the cases where data is suspected of being tampered with [@Schleunes2020; @Viglione2020]. The use of raw data allowed for issues to be identified, the absence of raw data not being available in other publications created difficulty in verifying the validity of the data.

Journals provide data papers in a formal publication explaining data that is archived. This is a familiar avenue for researchers to receive credit and citation. Writing a data paper can be very useful for researchers, however, data papers lack guidelines on how to structure data for effective reuse and sharing. Partner with Dryad to link archived data with research papers. Dryad will also provide professionals to curate your data, thereby improving usability.

One critical advantage of depositing data in a trusted repository (Section \@ref(avenues)) is that it comes with many benefits such as preservation, indexing,  discoverability and most importantly, citation with persistent identifier. Authors can take an extra step by including their own CITATION file to give credit to a paper where the data and associated methods are described in detail. When citing data, you should cite the preferred CITATION file in the first instance, else, if there is a data publication, (for example in _Scientific Data_), cite this, and if none of those apply, cite the repository where the data is deposited (e.g., a Zenodo repository).

@@

Data can be distributed alongside code using a format such as a package in R, Python, and Julia. A package or module structure provides an advantage as it can be "installed" and loaded just like code. Despite the advantages of sharing data as a software package, there are a few major disadvantages. Firstly, packaging data inside a software package makes the data availability language centric. Users who work in other languages are unlikely to download and export data out of a package. Secondly, data size can be a limitation depending on the repository (CRAN, the R repository, for example, limits package sizes to 5Mb although there are workarounds). Other fields, such as bioinformatics have a process of sharing data through repositories like Bioconductor, with larger data size. Thirdly, there is the cost of authors and data curators/custodians understanding the intricacies of package development in order to share data. Nevertheless, the frictionless approach to sharing data in software packages makes reuse substantially faster than retrieving and correctly parsing files from a repository. However, we recommend the software approach only as an additive to proper data curation.

Packaged data, and archival on venues like Zenodo are not mutually exclusive. First, archive data in a long-term data repository (Section \@ref(avenues)), before including it in a package. The act of packaging data in software can shave off hours required to prepare it for use. The canocical citation can be baked into the package's citation file.

@@

Small data has more options in how it is shared - you can bundle it with your analysis code (as in \@ref(rawdata)), or in software (as in \@ref(pkgs)). For medium data, consider a venue like Zenodo for archiving. Sharing medium data might require it to be broken into separate chunks for reading (e.g., `teaching-1.csv`, `teaching-2.csv`, etc.), and scripts provided to read it. Large data will require liaising with services from \@ref(avenues), or local research institute library to make special arrangements, to ensure there is some way to gain access, and also maintain a DOI.

@@

Although data sharing is widely mandated by funders, journals, and academic societies, the amount of usable data remains appallingly low. The mere act of depositing data does not make it reusable or useful to a potential consumer. Although data citations do not currently impact promotion and tenure packets, there is anecdotal evidence that papers sharing usable data accrue more citations [@Piwowar2007]. Over time, the more useful we make the data, the more likely incentive mechanisms will take hold.
